# Packages commonly used at Grattan

R can do a lot out of the box, but a range of [packages](#packages) make our lives easier.

Some packages - like the `tidyverse` collection of packages - are broadly popular among R users. Some - like the `grattantheme` package - are specific to Grattan Institute. Others - like the `readabs` package - are made by Grattan people, useful at Grattan, but also used outside of the Institute.

## The tidyverse! {#tidyverse}

The `tidyverse` is a [collection of related R packages](https://www.tidyverse.org/packages/) that all work well together. You install it in the [usual way](#install-packages):

```{r, eval = FALSE}
install.packages("tidyverse")
```

The main packages in the tidyverse include:

* *ggplot2* for making beautiful, customisable graphs
* *dplyr* for manipulating data frames
* *tidyr* for tidying your data
* *readr* for importing data from a broad range of formats
* *purrr* for functional programming
* *stringr* for manipulating strings of text

All these packages (and more!) will automatically be loaded for you when you run the command:

```{r library-tidyverse}
library(tidyverse)
```

There's no need to install or load the individual `tidyverse` packages - like `dplyr` - separately. Just install them all together, and load them with the single `library(tidyverse)` command. That way, you don't need to remember which functions come from `tidyr` and which from `dplyr` - they're all just `tidyverse` functions.

A range of other packages are installed on your machine as part of the `tidyverse.` These include:

* *readxl* for importing Excel spreadsheets into R
* *haven* for importing Stata, SAS and SPSS data
* *lubridate* for working with dates
* *rvest* for scraping websites

Although these packages are installed as part of the `tidyverse`, they aren't loaded automatically when you run `library(tidyverse)`. You'll need to load them individually, like:

```{r, eval = FALSE}
library(lubridate)
```

### Why do we use the tidyverse?

The `tidyverse` makes life easier! 

The core `tidyverse` packages, like `ggplot2`, `dplyr`, and `tidyr`, are extremely popular. They're probably the most popular 'dialect' of R. This means that any problem you encounter with the `tidyverse` will have been encountered many times before by other R users, so a solution will only be a Google search away.

The `tidyverse` packages are all designed to work well together, with a consisten underlying philosophy and design. They're designed to work with data frames[^The tidyverse works with 'tibbles', which are a tidyverse-specific variant of the data frame. Don't worry about the difference between tibbles and data frames.], a rectangular data object that will be familiar to spreadsheet users that is very intuitive and convenient for the sort of work we do at Grattan.

The creator of the `tidyverse`, Hadley Wickham, places great value on code that is expressive and comprehensible to humans. 

## Grattan-specific packages



# Writing reproducible documents with RMarkdown
