# Chart cookbook

This section takes you through a few often-used chart types. 


## Set up

```{r, include = FALSE}
# Set figure output to replicate type = "normal" 
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 8.74, # inches
                      fig.height = 5.71) # inches
```


```{r load_packages, message = FALSE}
library(tidyverse)
library(grattantheme)
library(ggrepel)
library(absmapsdata)
library(sf)
library(scales)
library(ggfittext)
```


```{r grattan_label}
# note: to be added to grattantheme; remove this when done
grattan_label <- function(..., size = 18) {

  .size = size / ggplot2::.pt
  
geom_label_repel(..., 
           fill = "white",
           label.padding = unit(0.1, "lines"), 
           label.size = 0,
           size = .size)
}
```


The `sa3_income` dataset will be used for all key examples in this chapter.^[From [ABS Employee income by occupation and sex, 2010-11 to 2015-16](https://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/6524.0.55.0022011-2016?OpenDocument)] It is a long dataset from the ABS that contains the median income and number of workers by Statistical Area 3, occupation and sex between 2010 and 2015.

```{r read_data}
sa3_income <- read_csv("data/sa3_income.csv") %>% 
  filter(!is.na(median_income),
         !is.na(average_income))

head(sa3_income)
```



## Bar charts

Bar charts are made with `geom_bar` or `geom_col`. Creating a bar chart will look something like this:


```{r bar_eg, eval = FALSE}
ggplot(data = <data>) + 
  geom_bar(aes(x = <xvar>, y = <yvar>),
     stat = <STAT>, 
     position = <POSITION>
  )
```


It has two key arguments: `stat` and `position`. 

First, `stat` defines what kind of _operation_ the function will do on the dataset before plotting. Some options are:

- `"count"`, the **default**: count the number of observations in a particular group, and plot that number. This is useful when you're using microdata. When this is the case, there is no need for a `y` aesthetic.
- `"sum"`: sum the values of the `y` aesthetic.
- `"identity"`: directly report the values of the `y` aesthetic. This is how PowerPoint and Excel charts work.

You can use **`geom_col`** instead, as a shortcut for `geom_bar(stat = "identity)`. 

Second, `position`, dictates how multiple bars occupying the same x-axis position will positioned. The options are:

- `"stack"`, the default: bars in the same group are stacked atop one another.
- `"dodge"`: bars in the same group are positioned next to one another.
- `"fill"`: bars in the same group are stacked and all fill to 100 per cent.

First, we'll take the `sa3_income` dataset and create a summary table of total workers and income by state, greater capital city area, professional work and gender. We'll use this `base_data` throughout this section.

```{r bar_basedata}
base_data <- sa3_income %>% 
  filter(!gender == "Persons") %>% 
  mutate(income = average_income * workers,
         prof = if_else(occupation %in% c("Professionals", "Managers"),
                        "Professional",
                        "Non-professional"),
         prof = factor(prof, levels = c("Professional", "Non-professional"))) %>% 
  group_by(state, gcc_name, prof, gender) %>% 
  summarise(workers = sum(workers),
            income = sum(income),
            average_income = income / workers)

```

Time to create a bar chart!^[Wooh! (Sorry, that's an abuse of footnotes.)]
You'll plot average incomes by state and gender, so first thing you'll do is **create the dataset that you want to plot**. 


```{r bar1_data}
data <- base_data %>% 
  group_by(state, gender) %>% 
  summarise(average_income = sum(income) / sum(workers))

data
```

Looks good: you have one observation (row) for each state $\times$ gender group you want to plot, and a value for their average income. Put `state` on the x-axis, `average_income` on the y-axis, and split gender by fill-colour (`fill`).^[The aesthetic `fill` represents the 'fill' colour -- the colour that fills the bars in your chart. The `colour` aesthetic controls the colours of the _lines_.] 
And because you have the _actual_ values you want to plot from your `data` object, tell `geom_bar` to 'just use the number provided' with `stat = "identity"`:

```{r bar1a}
data %>% 
  ggplot(aes(x = state,          # x-axis is state
             y = average_income, # y-axis is income
             fill = gender)) +      # fill colour is gender
  geom_bar(stat = "identity") +  # "just use the number that's there"
  theme_grattan() +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "")
```

Before addressing the accidental stacking, note that there is a short-cut-geom for `geom_bar(stat = "identity")`: `geom_col`. Using that instead, without the need for the `stat` argument gives the same result:

```{r bar_col}
data %>% 
  ggplot(aes(x = state,        
             y = average_income,
             fill = gender)) +     
  geom_col() +                 # no need for stat = "identity"
  theme_grattan() +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "")
```

The `fill` aesthetic creates two series -- one for women, one for men -- and `geom_col` stacks them by default. You can change this behaviour with the `position` argument:  

```{r bar1b}
data %>% 
  ggplot(aes(x = state,         
             y = average_income,
             fill = gender)) +    
  geom_col(position = "dodge") + # dodge instead of stack the series
  theme_grattan() +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "")
```


You can also **order** the groups in your chart by a variable. If you want to order states by population, use `reorder` inside `aes`:

```{r bar2}
data %>% 
  ggplot(aes(x = reorder(state, -average_income), # change the order of state
             y = average_income,
             fill = gender)) +
  geom_col(position = "dodge") +
  theme_grattan() +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "")
```

To flip the chart -- a useful move when you have long labels -- add `coord_flipped` (ie 'flip coordinates') and tell `theme_grattan` that the plot is flipped using `flipped = TRUE`. 

However, while the _coordinates_ have been flipped, the underlying data hasn't. If you want to refer to the `average_income` axis, which now lies horizontally, you would still refer to the `y` axis (eg `grattan_y_continuous` still refers to your `y` aesthetic, `average_income`). 

```{r bar4}
data %>% 
  ggplot(aes(x = reorder(state, -average_income),
             y = average_income,
             fill = gender)) +
  geom_col(position = "dodge") +
  theme_grattan(flipped = TRUE) +   # tell theme_grattan of flipping
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "") + 
  coord_flip() # flip the chart
```

### Labelling bar charts

Labelling charts in the Grattan style means labelling _one_ of each series and letting the reader connect the dots. 

First, create a base chart that we'll use for the next few charts:

```{r}
bar <- data %>% 
  ggplot(aes(x = reorder(state, -average_income),
             y = average_income,
             fill = gender)) +
  geom_col(position = position_dodge()) +
  theme_grattan(flipped = TRUE) +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "") + 
  coord_flip()

bar
```

You can use `grattan_label` to label your charts. This function is a 'wrapper' around `geom_label_repel` that has settings that we tend to like: white background with a thin margin, 18-point font, and no border. It takes the standard arguments of `geom_label_repel`. 

Section \@ref(adding-labels) shows how labels are treated like data points: they need to know where to go (`x` and `y`) and what to show (`label`). But if you provide _every point_ to your labelling `geom`, it will plot every label:

```{r}
bar +
  grattan_label(aes(colour = gender,  # colour the text according to gender
                    label = gender),  # label the text according to gender
            position = position_dodge(width = 1),  # position dodge with width 1
            hjust = -0.1) +  # horizontally align the label so its outside the bar
  grattan_colour_manual(2)   # define colour as two grattan colours
```

To just label _one_ of the plots -- ie the first one, Tasmania in this case -- we need to tell `grattan_label`. The easiest way to do this is by **creating a label dataset beforehand**, like `label_gender` below. This just includes the observations you want to label:

```{r}
label_gender <- data %>% 
  filter(state == "Tas")  # just want Tasmania observations

bar +  
  grattan_label(data = label_gender,  # supply the new dataset
                aes(colour = gender,
                    label = gender), 
                position = position_dodge(width = 1), 
                hjust = -0.1) + 
  grattan_colour_manual(2)

```


Labelling a vertical bar chart is much the same, but you might find yourself with less horizontal space. After creating a new base chart, our previous technique is a bit off:

```{r}
# Create vertical bar chart
vbar <- data %>% 
  ggplot(aes(x = reorder(state, -average_income),
             y = average_income,
             fill = gender)) +
  geom_col(position = position_dodge()) +
  theme_grattan() +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "")

# Create label dataset
label_gender <- data %>% 
  filter(state == "ACT")  # just want ACT observations

# Label chart 
vbar + 
  grattan_label(data = label_gender,  # supply the new dataset
              aes(colour = gender,
                  label = gender), 
              position = position_dodge(width = 1), 
              hjust = 0.5,  # centre horizontally
              vjust = -.1) + # nudged up a bit
  grattan_colour_manual(2)

```

It might be best to label the bars in empty space _above_ the bars. First, expand the plot area by adding an `expand_top` argument to `grattan_y_continuous`:

```{r, message = FALSE}
vbar + 
  grattan_y_continuous(labels = dollar,
                       expand_top = 0.2)
```

That gives you a bit of breathing room. Use the same `label_gender` dataset to place labels above the first state (ACT). Like before, `grattan_label` is used to plot the labels -- given a dataset and the right aesthetics -- but you'll need to tweak some of the settings to get it looking right.

By using `nudge_y = Inf`, you're pushing the labels all the way to the top of the plot. Then, `direction = "x"` says 'only move the labels horizontally to make sure they're not overlapping'. Finally, `segment.colour = NA` turns off the line that connects the label to the data point (you can leave this out and see if you like the look of it!).


```{r, message = FALSE}
# Create label dataset
label_gender <- data %>% 
  filter(state == "ACT")  

# Plot with labels
vbar + 
  grattan_y_continuous(labels = dollar,
                     expand_top = 0.07) + 
  grattan_label(data = label_gender,
              aes(colour = gender,
                  label = gender), 
              nudge_y = Inf, # move to the top of the plotting area
              direction = "x", # automatically align along the x-axis
              segment.colour = NA) +  # don't add a line to the data point
  grattan_colour_manual(2)

```


There is an alternative approach that can be used in situations that require only a couple of labels. The `annotate` function will add text to a plot in the place you tell it. It doesn't require a new dataset, and is more like manually labelling a plot in Powerpoint. 

The first argument of annotate is what kind of _thing_ it is: `"text"`. Then set the label text you'd like, eg `"label = Men"`, and where it should go with `x` and `y`. Then set the `colour`, and fiddle with `hjust` (horizontal alignment) to get your desired placement.

The outcome is close to what we had with `grattan_label`:

```{r, message = FALSE}
vbar + 
  grattan_y_continuous(labels = dollar,
                       expand_top = 0.1) + 
  annotate("text", label = "Men",  # Male label
           x = "ACT", y = 75e3, 
           colour = grattan_orange,
           hjust = 1) + 
  annotate("text", label = "Women", # Female label
           x = "NT", y = 75e3, 
           colour = grattan_red,
           hjust = 1)

```


### Facetting bar charts {#facet-bar}

'Facetting' a chart means you create a separate plot for each group. It's particularly useful in showing differences between groups.

You can 'facet' bar charts -- and any other chart type -- with the `facet_grid` or `facet_wrap` commands. The `fadcet_wrap` tends to give you more control over label placement, so let's start with that. 


First, create a summary dataset of average income by professional employment, state and gender:

```{r}
data <- base_data %>% 
  group_by(prof, state, gender) %>% 
  summarise(average_income = sum(income) / sum(workers))

head(data)
```

Then plot a bar chart, similar to the one you made before, but add `facet_wrap` to the chain:

```{r}
data %>% 
  ggplot(aes(x = reorder(state, average_income),         
             y = average_income,
             fill = gender)) +  
  geom_col(position = position_dodge()) + 
  theme_grattan(flipped = TRUE) +
  grattan_y_continuous(labels = dollar) +
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "") + 
  coord_flip() + 
  facet_wrap(prof ~ .) # facet 'prof' around nothing else '.'
```

Cool! Now you just have to tweak some settings to get the plot looking right on the page:

- Add a black line along `average_income = 0` to distinguish the plots. 
- Define your 'breaks' (the spacing of the axis gridlines and labels) in `grattan_y_continuous`. 
- Add some additional space -- `10mm` -- between the two facet plots with `panel.spacing = unit(10, "mm")` in the `theme` parameter.^[The `unit` function takes two arguments: the number of the unit, then the unit itself (eg "mm" for millimetre, or "cm" for centremetre, etc.)] 
- Left-align the facet titles with `strip.text = element_text(hjust = 0)`.^[The function `element_text` is useful for adjusting the look of text in a plot. With it, you can say you want the text to be red, `colour = "red"`, or MASSIVE, `size = 100`, or **bold**, `face = "bold"`.] 
- We're a bit tight for space along the bottom, so get rid of the "$" and replace with "comma", then add that information in the `y` label.


```{r}
facet_bar <- data %>% 
  ggplot(aes(x = reorder(state, average_income),         
             y = average_income,
             fill = gender)) +  
  geom_col(position = position_dodge()) + 
  geom_hline(yintercept = 0) + # add a black base line for each facet
  theme_grattan(flipped = TRUE) +
  grattan_y_continuous(labels = comma, # change to comma
                       breaks = c(0, 25e3, 50e3, 75e3)) + # define your breaks
  grattan_fill_manual(2) + 
  labs(x = "",
       y = "Annual income, $") + # add $ info in y-label
  coord_flip() + 
  theme(panel.spacing = unit(10, "mm"), # add some space between the plots
        strip.text = element_text(hjust = 0)) + # left-align facet titles
  facet_wrap(prof ~ .)

facet_bar
```

**Finally**, we'll add the women/men labels by creating a dataset with the labels we want and giving that to `grattan_label`. 


```{r}
label_data <- data %>% 
  filter(state == "ACT",
         prof == "Non-professional")



facet_bar +
  grattan_label(data = label_data,  # supply the new dataset
                aes(colour = gender,
                    label = gender), 
                position = position_dodge(width = 1), 
                hjust = -.1) + 
  grattan_colour_manual(2) # define the colour scale

```



## Line charts

A line chart has one key aesthetic: `group`. This tells `ggplot` how to connect individual lines. 

First, we'll take the `sa3_income` dataset and create a summary table of total workers and income by year, state, greater capital city area, professional work and gender. We'll use this `base_data` throughout this section.


```{r line1, message = FALSE}

```

You can also add dots for each year by layering `geom_point` on top of `geom_line`:

```{r line2, message = FALSE}

```

If you wanted to show each state individually, you could **facet** your chart so that a separate plot was produced for each state:

```{r line3, message = FALSE}

```

To tidy this up, we can: 

  1. shorten the years to be "13", "14", etc instead of "2013", "2014", etc (via the `x` aesthetic)
  1. shorten the y-axis labels to "millions" (via the `y` aesthetic)
  1. add a black horizontal line at the bottom of each facet
  1. give the facets a bit of room by adjusting `panel.spacing`
  1. define our own x-axis label breaks to just show `13`, `15` and `17`



```{r line4, message = FALSE}

```



## Scatter plots

Scatter plots require `x` and `y` aesthetics. These can then be coloured and faceted.

First, create a dataset that we'll use for scatter plots. Take the `population_table` dataset and transform it to have one variable for population in 2013, and another for population in 2018:

```{r scatter_data, message = FALSE}


```

 
Then plot it  
 
```{r scatter, message = FALSE}

```


```{r scatter_rescale, message = FALSE}

```


It looks like the areas with the largest population grew the most between 2013 and 2018. To explore the relationship further, you can add a line-of-best-fit with `geom_smooth`:

```{r scatter_smooth, message = FALSE}

```


You could colour-code positive and negative changes from within the `geom_point` aesthetic. Making a change there won't pass through to the `geom_smooth` aesthetic, so your line-of-best-fit will apply to all data points.

```{r scatter_colour, message = FALSE}

```


Like the charts above, you could facet this by state to see if there were any interesting patterns. We'll filter out ACT and NT because they only have one and two data points (SA4s) in them, respectively.

```{r scatter_facet, message = FALSE, warning = FALSE}

```


## Distributions

`geom_histogram`
`geom_density`

`ggridges::`


## Maps

### `sf` objects
[what is]

### Using `absmapsdata`

The `absmapsdata` contains compressed, and tidied `sf` objects containing geometric information about ABS data structures. The included objects are:

  - Statistical Area 1 2011 and 2016: `sa12011` or `sa12016`
  - Statistical Area 2 2011 and 2016: `sa22011` or `sa22016`
  - Statistical Area 3 2011 and 2016: `sa32011` or `sa32016`
  - Statistical Area 4 2011 and 2016: `sa42011` or `sa42016`
  - Greater Capital Cities 2011 and 2016: `gcc2011` or `gcc2016`
  - Remoteness Areas 2011 and 2016: `ra2011` or `ra2016`
  - State 2011 and 2016: `state2011` or `state2016`
  - Commonwealth Electoral Divisions 2018: `ced2018`
  - State Electoral Divisions 2018:`sed2018`
  - Local Government Areas 2016 and 2018: `lga2016` or `lga2018`
  - Postcodes 2016: `postcodes2016`

You can install the package from Github:

```{r install_absmapsdata, eval = FALSE}
remotes::install_github("wfmackey/absmapsdata")
library(absmapsdata)
```

You will also need the `sf` package installed to handle the `sf` objects:

```{r install_sf, eval = FALSE}
install.packages("sf")
library(sf)
```



### Making choropleth maps

Choropleth maps break an area into 'bits', and colours each 'bit' according to a variable.

SA4 is the largest non-state statistical area in the ABS ASGS standard. 

You can join the `sf` objects from `absmapsdata` to your dataset using `left_join`. The variable names might be different -- eg `sa4_name` compared to `sa4_name_2016` -- so use the `by` function to match them.

```{r make_map_data, message = FALSE}

```

You then plot a map like you would any other `ggplot`: provide your data, then choose your `aes` and your `geom`. For maps with `sf` objects, the **key aesthetic** is `geometry = geometry`, and the **key geom** is `geom_sf`.

The argument `lwd` controls the line width of area borders.

Note that RStudio takes a long time to render a map in the 


```{r map1}

```

Showing all of Australia on a single map is difficult: there are enormous areas that are home to few people which dominate the space. Showing individual states or capital city areas can sometimes be useful. 

To do this, filter the `map_data` object: 

```{r map_filter}

```


#### Adding labels to maps

You can add labels to choropleth maps with the standard `geom_text` or `geom_label`. Because it is likely that some labels will overlap, `ggrepel::geom_text_repel` or `ggrepel::geom_label_repel` is usually the better option.

To use `geom_(text|label)_repel`, you need to tell `ggrepel` where in 



```{r map_label, warning = FALSE, eval = FALSE}
map <- map_data %>% 
        filter(state == "Vic") %>% 
        ggplot(aes(geometry = geometry)) +
        geom_sf(aes(fill = pop_change),
                lwd = .1,
                colour = "black") +
        theme_void() +
        grattan_fill_manual(discrete = FALSE, 
                            palette = "diverging",
                            limits = c(-20, 20),
                            breaks = seq(-20, 20, 10)) +
  geom_label_repel(aes(label = sa4_name),
                  stat = "sf_coordinates", nudge_x = 1000, segment.alpha = .5,
                  size = 4, 
                  direction = "y",
                  label.size = 0, 
                  label.padding = unit(0.1, "lines"),
                  colour = "grey50",
                  segment.color = "grey50") + 
  scale_y_continuous(expand = expand_scale(mult = c(0, .2))) + 
  theme(legend.position = "top") + 
  labs(fill = "Population \nchange")

map
```


## Creating simple interactive graphs with `plotly`

`plotly::ggplotly()`


